Hydra Battles and Cie

### 2020-12-14 doc/hydras.tex
correction of typos

@@

\subsection{Contributions}
Many thanks to Yves Bertot, 
\'Evelyne Contejean,  Florian Hatat and Pascal Manoury  for their contribution. 
Yves made nice optimizations  to algorithms presented in Chapter~\ref{chapter-powers}.
\'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

@@

@@

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to Yves Bertot, David Ilcinkas, Karl Palmskog, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on  oral presentations of this work.

Many thanks also to the Coq development team, Yves Bertot, and members of the \emph{Coq Club} for interesting discussions about the \coq{} system and the Calculus of Inductive Constructions.

The author of the present document wishes to express his gratitude to the late Patrick Dehornoy, whose talk  was determinant for our desire to work on this topic.

I owe my interest in discrete mathematics and their relation to formal proofs and functional programming  to Srecko Brlek.  Equally, there is W. H. Burge's book ``\emph{Recursive Programming Techniques}'' ~\cite{burge} which was a great  source of inspiration.

@@

>>>>

\section{Generalities}

@@

@@

In general, we do not include full proof scripts in this document. The only exceptions are very short proofs (\emph{e.g.} proofs by computation, or by application of automatic tactics). Likewise, we may display only the important steps on a long interactive proof, for instance, in the following lemma (\vref{lemma:L-2_6-1}):

\begin{Coqsrc}
Lemma Lemma2_6_1 (alpha : T1) :  
  nf alpha -> forall beta,  beta t1< alpha  ->
  {n:nat | const_pathS n alpha beta}.
Proof.
  transfinite_induction alpha.
  (* ... *)
\end{Coqsrc}

The reader may consult the full proof scripts with Proof General or CoqIDE, for instance.

@@

@@

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to Yves Bertot, \'Evelyne Contejean,  Florian Hatat,  David Ilcinkas, 
Pascal Manoury,  Karl Palmskog, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on  oral presentations of this work.

Many thanks also to the Coq development team, Yves Bertot, and the members of the \emph{Coq Club} for interesting discussions about the \coq{} system and the Calculus of Inductive Constructions.

The author of the present document wishes to express his gratitude to the late Patrick Dehornoy, whose talk  was determinant for our desire to work on this topic.
I owe my interest in discrete mathematics and their relation to formal proofs and functional programming  to Srecko Brlek.  Equally, there is W. H. Burge's book ``\emph{Recursive Programming Techniques}'' ~\cite{burge} which was a great  source of inspiration.

\subsection{Contributions}

Yves Bertot made nice optimizations  to algorithms presented in Chapter~\ref{chapter-powers}.
\'Evelyne Contejean contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

### 2020-12-14 doc/part-hydras.tex
correction of typos

@@

Thus, any result about \texttt{steps} will be applicable to standard battles.
Using the predicate \texttt{steps} our study of the length of the considered battle
can  be decomposed into three parts:

@@

The theorems \texttt{Alive\_free} and \texttt{Alive\_standard} of the module 
\url{../theories/html/hydras.Hydra.Hydra_Theorems.html} show that the classes \texttt{free} and \texttt{standard} satisfy this property.

@@

\begin{Coqsrc}
Definition  next_round_dec n :
 forall h , (h = head) + {h' : Hydra & {R1 h h'} + {R2 n h  h'}}.
\end{Coqsrc}

@@

In order to get a contradiction, it suffices to  prove the inequality
\texttt{m big\_h <= m small\_h}, i.e.,  \texttt{m big\_h <= m (iota (m big\_h))}.

@@

Let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a small sequence of ordinal numbers, which extends the sequence of natural numbers.

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots,\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots,\\
&\textcolor{red}{\omega^\omega},\ldots, \textcolor{red}{\omega^\omega+\omega^7\times 8},\ldots,\textcolor{red}{\omega^\omega\times 2},\omega^\omega\times 2+1, \ldots,\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^{\epsilon_0)}}, \epsilon_0+1, \epsilon_0+2, \epsilon_0+3, \ldots,\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots,\\
&\ldots
  \end{align*}   
  \end{minipage}}

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for  infinite sequences of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

Having the same order type is an equivalence relation between well-ordered sets. Ordinal numbers (in short \emph{ordinals}) are descriptions (\emph{names}) of the equivalence classes.
For instance, the order type of $(\mathbb{N},<)$ is associated with the ordinal called  $\omega$, and the order we considered on 
the disjoint union of $\mathbb{N}$ and itself is named $\omega+\omega$.

@@

\begin{remark}
The infix notations \texttt{o<} and \texttt{o<=} were defined in order to make apparent the distinction between the various notation scopes that may co-exist in a same statement. So the infix \texttt{<} and \texttt{<=} are reserved to the natural numbers. In the mathematical formulas, we still use $<$ and $\leq$ for comparing ordinals.
\end{remark}

@@

\section{Example: The Ordinal \texorpdfstring{$\omega$}{omega}}

@@

\section{Sum of  Ordinal Notations}

@@

\noindent
\emph{From Library~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Relations.Relation\_Operators}}.

@@

Let us look again at our implementation of $\omega+\omega$. We can distinguish between the three kinds of ordinals seen in Fig~\ref{fig:ordinal-sequence}:

\begin{itemize}
\item The least ordinal, \texttt{(inl 0)}, also written \texttt{(fin 0)}.
\item The limit ordinal $\omega$.
\item The successor ordinals, either of the form \texttt{(inl (S $i$))} or \texttt{(inr (S $i$))}
\end{itemize}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Prelude.Restriction.html}{Ordinals.Prelude.Restriction}}

@@

\begin{Coqsrc}
Lemma oplus_comm_0 : forall alpha, nf alpha ->
     forall a b,  nf a -> nf b ->
                  lt a alpha ->
                  lt b alpha ->
                  a o+ b = b o+ a.
 Proof with eauto with T1.
    intros alpha Halph; transfinite_induction alpha.
(* rest of proof omitted *)  
\end{Coqsrc}

@@

On may ask whether the preceding results are not too restrictive, since they 
refer to a particular data type \texttt{T1}.
In fact, our representation of ordinals strictly less than 
 $\epsilon_0$ is faithful to their mathematical definition, at least 
Kurt Schütte's~\cite{schutte}, as proved in Chapter~\vref{chap:schutte}.
(please see also \href{../theories/html/hydras.Schutte.Correctness_E0.html}{the module \texttt{Ordinals.Schutte.Correctness\_E0}}).

@@

This chapter describes an adaptation to \coq{} of Schütte's axiomatization. 
 Unlike the rest of our libraries, our library
\href{../theories/html/hydras.Schutte.Schutte.html}{Ordinals.Schutte}
is not constructive, and relies on several axioms.

>>>>

The following property allows us to build battles out of smaller ones.

\begin{Coqsrc}
Lemma battle_trans {b:Battle} :
  forall i h j h', battle b i h j h' ->
        forall k h0,  battle b k h0 i h -> battle b k h0 j h'.
\end{Coqsrc}

@@

Thus, any result about \texttt{steps} will be applicable to standard battles.
Using the predicate \texttt{steps},  our study of the length of the considered battle
can  be decomposed into three parts:

@@

The theorems \texttt{Alive\_free} and \texttt{Alive\_standard} of the module 
\href{../theories/html/hydras.Hydra.Hydra_Theorems.html}{Hydra.Hydra\_Theorems} show that the classes \texttt{free} and \texttt{standard} satisfy this property.

@@

\begin{Coqsrc}
Definition  next_round_dec n :
 forall h, (h = head) + {h' : Hydra & {R1 h h'} + {R2 n h  h'}}.
\end{Coqsrc}

@@

In order to get a contradiction, it suffices to  prove the inequality
$m(\texttt{big\_h}) \leq m(\texttt{small\_h})$ i.e.,  $m(\texttt{big\_h}\leq m(\iota (m(\texttt{big\_h})))$.

@@

Let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a few items of a  sequence of ordinal numbers, which extends the sequence of natural numbers.

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots\\
&\textcolor{red}{\omega^\omega},\ldots, \textcolor{red}{\omega^\omega+\omega^7\times 8},\ldots,\textcolor{red}{\omega^\omega\times 2},\omega^\omega\times 2+1, \ldots\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^{\epsilon_0)}}, \epsilon_0+1, \epsilon_0+2, \epsilon_0+3, \ldots\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots\\
&\ldots
  \end{align*}   
  \end{minipage}}

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for  infinite sequences of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, somewhere between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

Having the same order type is an equivalence relation between well-ordered sets. Ordinal numbers (in short: \emph{ordinals})  are descriptions (\emph{names}) of the equivalence classes.
For instance, the order type of $(\mathbb{N},<)$ is associated with the ordinal called  $\omega$, and the order we considered on 
the disjoint union of $\mathbb{N}$ and itself is named $\omega+\omega$.

@@

\begin{remark}
The infix notations \texttt{o<} and \texttt{o<=} were defined in order to make apparent the distinction between the various notation scopes that may co-exist in a same statement. So the infix \texttt{<} and \texttt{<=} are reserved to the natural numbers. In the mathematical formulas, however, we still use $<$ and $\leq$ for comparing ordinals.
\end{remark}

@@

\section{Example: the Ordinal \texorpdfstring{$\omega$}{omega}}

@@

\section{Sum of  two Ordinal Notations}

@@

\vspace{4pt}
\noindent
\emph{From Library~\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Relations.Relation_Operators.html}{Relations.Relation\_Operators}}.

@@

Let us look again at our implementation of $\omega+\omega$. We can classify its elements into three categories:

\begin{itemize}
\item The least ordinal, \texttt{(inl 0)}, also known as  \texttt{(fin 0)}.
\item The limit ordinal $\omega$.
\item The successor ordinals, either of the form \texttt{(inl (S $i$))} or \texttt{(inr (S $i$))}
\end{itemize}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Prelude.Restriction.html}{hydras.Prelude.Restriction}}

@@

\begin{Coqsrc}
Lemma oplus_comm_0 : forall alpha, nf alpha ->
     forall a b,  nf a -> nf b ->
                  lt a alpha ->
                  lt b alpha ->
                  a o+ b = b o+ a.
 Proof with eauto with T1.
    intros alpha Halpha; transfinite_induction alpha.
(* rest of proof omitted *)  
\end{Coqsrc}

@@

On may ask whether the preceding results are not too restrictive, since they 
refer to a particular data type \texttt{T1}.
In fact, our representation of ordinals strictly less than 
 $\epsilon_0$ is faithful to their mathematical definition, at least 
Kurt Schütte's~\cite{schutte}, as proved in Chapter~\vref{chap:schutte}.
(please see also \href{../theories/html/hydras.Schutte.Correctness_E0.html}{the module \texttt{hydras.Schutte.Correctness\_E0}}).

@@

This chapter describes an adaptation to \coq{} of Schütte's axiomatization. 
 Unlike the rest of our libraries, our library
\href{../theories/html/hydras.Schutte.Schutte.html}{hydras.Schutte}
is not constructive, and relies on several axioms.
